---
title: "Summary Measures, Statistical Relationships, and Distributions"
format: html
editor_options: 
  chunk_output_type: console
---

Import packages that we'll use:

```{r}
library(tidyverse)
```

# Statistical Relationships

As we explore the relationship between different types of variables, we often want to test whether those relationships are significant: are they real or likely to hold with other data, or are they part of the expected variability and noise in the data?

The type of statistical test we use depends on whether we're working with continuous or categorical variables.

::: callout-note
## This is not a statistics workshop

All of the statistical measures and tests we're covering have assumptions about the data and other requirements to produce valid output. Here, we're focused on how to run the tests in R, not how to ensure we're making correct statistical inferences. Please consult your stats book or a statistician for guidance on actual statistical inference and analysis decisions.
:::

## Correlation: Two Continuous

One way to look at the relationship between two continuous variables is the correlation. For example, height and weight:

```{r}
ggplot(healthdata, aes(cholesterol, weight)) +
  geom_point()
```

Correlation looks at whether there is a linear relationship.  `cor()` takes two vectors.

```{r}
cor(healthdata$cholesterol, healthdata$weight)
```

`NA` again!

The `NA` options for `cor()` are slightly different than for `mean()` or other similar functions.  The `NA` options are different because `cor()` can also be used with matrices of numeric values.  When computing the correlation between two variables, the option to use is:

```{r}
cor(healthdata$cholesterol, healthdata$weight, use = "pairwise")
```

To determine whether a correlation is statistically significant, we can use `cor.test()`

```{r}
cor.test(healthdata$cholesterol, healthdata$weight, use = "pairwise")
```

In addition to computing the correlation coefficient, which ranges between -1 and 1, it also computes a t-test (more on these below) where the null hypothesis is that there is no correlation between the variables.  p-values < 0.05 are often considered to be statistically significant, meaning that there is less than a 5% chance of seeing the given correlation statistic if there was truly no relationship between the two variables.

The result is affected by how many observations are included.

## TRY IT

Compute the correlation between `bp_sys1` and `bp_dia1.`  Is this correlation statistically different from 0?

```{r}

```


### Tidyverse

You can compute a correlation within a Tidyverse workflow:

```{r}
healthdata %>%
  summarize(correlation = cor(height, weight, use="pairwise"))
```



## t-test: Categorical and Continuous

Is the average height of female participants different than that of male participants?  This is an example of investigating the relationship between a continuous variable (height) and a categorical variable (gender). 

To do this, we can use a t-test to see how the mean differs across two groups.

The easiest way to specify that we want to compare the average value of a variable between two groups is to use R's formula syntax, which includes a `~`.  `~` is read as "as a function of".  So to test whether average height differs as a function of gender:

```{r}
t.test(healthdata$height ~ healthdata$gender)
```

The output includes the mean values for each group, a confidence interval on the *difference* between the two means, and the test statistic and p-value.  p-values < 0.05 are often considered to be statistically significant, meaning that there is less than a 5% chance of seeing the given test statistic (t value) if there was truly no difference in means between the two groups.

If the continuous variable has more than 2 groups, you will need a test or model other than a t-test, such as ANOVA.  

## TRY IT

Does average pulse differ by gender?

```{r}

```


## t-test: Accessing Results

If you need the values from the t-test results, instead of copying them from the output in the console, you can extract them from the result of the test.  Save the test result to a variable, and then you can access its components:

```{r}
t1 <- t.test(healthdata$height ~ healthdata$gender)
names(t1)  # works for a variety of objects besides data frames
t1$statistic
t1$p.value
```


## t-test: Other Scenarios

t-tests can also be used in other situations, such as comparing pre and post measures for participants in an experiment or testing whether a single value, such as the mean of a variable, is different from 0 (or another value).

### Single Variable t-test

Instead of comparing the mean of two groups of observations, we can compare whether the mean is different than a specific value.  The specific value of interest is often 0, but we can compare the mean to any value.

The average height of people in the US is about 168 cm.  Does the average height of people in this data set differ from that?

We can specify a comparison value with the `mu` argument, which defaults to 0.

```{r}
t.test(healthdata$height, mu = 168)
```



### Paired t-test

A paired t-test is used when you have more than one measurement per observation.  Instead of comparing averages across two groups, we instead test whether the difference between measurement 1 and measurement 2, averaged across observations (people), is different than 0 (or another value).

With our dataset, we can ask: is the first BP measurement different than the second?  We'll compare bp_sys1 and bp_sys2, and indicate that the observations are paired:

```{r}
t.test(healthdata$bp_sys1, healthdata$bp_sys2, paired = TRUE)
```


## Chi-squared: Two Categorical

To test whether there is a relationship between two categorical variables, we can use a chi-squared test.  

For example, do men and women have different education levels?

```{r}
table(healthdata$gender, healthdata$education)
prop.table(table(healthdata$gender, healthdata$education), 
           margin = 1)  # 1 = rows
```

A chi-squared test doesn't tell us what the relationship is, just whether there is a difference in proportions across categories between different groups.

```{r}
chisq.test(healthdata$gender, healthdata$education)
```

## TRY IT

Is there a relationship between gender and work_status?  Make a table first to look at the relationship, and then run a chi-squared test.

```{r}

```

# Linear Regression: Continuous Dependent Variable

Linear regression is used to determine the relationship between one or more independent or explanatory variables and a continuous dependent or outcome variable.  

We're going to work with pulse as the dependent variable.  If we just had one predictor, age, then linear regression would fit a line to the data that minimizes the sum of the squared errors (the squared distance between each point and the fitted line)

```{r}
ggplot(healthdata, aes(x=age, y=pulse)) +
  geom_point(alpha = .3)  # transparency to partially address overlapping points
```

The linear regression line:

```{r}
ggplot(healthdata, aes(x=age, y=pulse)) +
  geom_point(alpha = .1) +
  geom_smooth(method = "lm")  # "lm" corresponds to the linear model lm() function
```

We want to do more than plot the line - we want to know the slope, intercept, and other measures of the model fit.


## Formula Syntax: Basics

To compute the linear regression model, we use the `lm`() function (lm = linear model).  We specify the relationship between the variables with the formula syntax, which at the basic level is `y ~ x`.  The y is the dependent or outcome variable, and it is being modeled as a function of one or more x (independent, predictor, explanatory) variables

```{r}
lm(pulse ~ age, data=healthdata)
```

The basic output includes the intercept and slope of the line.  

## Summary

We can get additional information about the model fit with the `summary()` function.  This is the same function that can be used to get summary statistics about a data frame or variable.  The function takes different actions depending on what type of input it is given.

We give the output of `lm()` to summary.  A few options for how to do this:

```{r}
# pipe
lm(pulse ~ age, data=healthdata)  %>%  
  summary()
```

```{r, eval=FALSE}
# nested
summary(lm(pulse ~ age, data=healthdata))

# save the result in a variable
reg1 <- lm(pulse ~ age, data=healthdata)
summary(reg1)
```

The output of `summary()` includes the standard error on each coefficient estimate, and it computes a t-test for each to determine the likelihood that the coefficient estimate would take the given value if the true value of the coefficient was 0.  We get "stars" summarizing which coefficients are statistically different from 0 at common significance levels. It also include the R-squared measure, which is a measure of how much variance in the dependent variable is explained by the model.  It ranges between 0 and 1.  

The output of `summary()` also let's us know that 184 observations were dropped from the model because they had a missing value for at least one of the variables included in the model.  Observations (rows) must have non-missing values for all variables in the model to be included in the estimation.  

## TRY IT

Run a linear regression model explaining weight with height.  Use `summary()` to see detailed output.

```{r}

```

